Curve counting in genus one: elliptic singularities {\it \&} relative geometry

### 2019-06-28 EllipticSingularitiesSplitting.tex
improved description of D from Dtilde

\subsubsection{$\Dcal$ from $\widetilde\Dcal$}\label{subsection D from Dtilde} On $\widetilde\Dcal$ there is a tautological vector bundle arising from the modular description of $\widetilde\VZ_{1,\alpha}(\PP^N|H,d)$, namely:
\begin{equation*} F = \OO(\delta)^{\oplus r}. \end{equation*}

\begin{lemma}
The fibre of $F$ over a moduli point can be identified with
\begin{equation*} \left( \bigoplus_{i=1}^r \TT_{p_i} C_i^\prime \right) \otimes \TT_{C_0} \end{equation*}
where $C_i^\prime$ is a component of $C_i$ which lies on the radius, $p_i$ is the adjacent node which points towards the core, and $\TT_{C_0}$ is the universal tangent line bundle.
\end{lemma}

\begin{proof}
Choose a component $C_i^\prime$ of the curve which lies on the radius. Then $\OO(\delta)$ is equal to the tensor product of tangent line bundles along the path of edges connecting $C_i^\prime$ to the core. However, this tensor product of line bundles is telescoping: any vertex which appears in the middle of the path will have an incoming and outgoing node, and since the vertex corresponds to a smooth rational curve these tangent spaces are naturally dual to each other~\cite{VZ}. Standard manipulations lead to the result.

%Note that the final term on the right hand side is not given by $\TT_{q_i} C_0$ (where $q_i$ is the appropriate splitting node) but rather the tangent space to the core. 
\end{proof}

We said that $C_i^\prime$ is \emph{a} component of $C_i$ which lies on the radius. There may be more than one, but if this is the case then their tangent spaces are naturally identified.

The following construction allows us to give a canonical description of $\Dcal$ inside $\widetilde\Dcal$. Over $\widetilde\Dcal$ there is a natural section of the projective bundle $\PP(F) \to \widetilde\Dcal$ (obtained by projectivising the diagonal embedding). We may then pullback to $\widetilde\Dcal$ the map of vector bundles
\begin{equation}\label{eq:factorisation} \OO_{\PP(F)}(-1)(-\TT_{C_0}) \xrightarrow{\text{can}(-\TT_{C_0})} \pi^\st F(-\TT_{C_0}) \xrightarrow{\pi^*\sum\operatorname{d}f} \pi^\st(\ev_q^\st \TT_H),\end{equation}\footnote{We may assume the latter factors through $\ev_q^\st \TT_H\subseteq\ev_q^\st \TT_{\PP^N}$ by virtue of Lemma \ref{lemma type C0 combinatorial types}.}
whose vanishing tautologically expresses the condition that the stable map $f$ descend to the elliptic singularity associated to a point of $\widetilde\Dcal$. Thus, the class of $\Dcal$ can be represented via an Euler class computation. Note that the vector bundle $F$ can be explicitly realised by twisting the pullback of $\bigoplus_{i=1}^rT_{q_i}C_i$ from $\Ecal$ with a number of exceptional boundary corrections.

\begin{comment}
We obtain a natural morphism
\begin{equation*} \sum \operatorname{d}\!f \colon F(-\TT_{C_0}) \to \ev_q^\st \TT_H \end{equation*}
given by summing the derivatives at \emph{all} the inward-pointing nodes on the radius. {\bf \color{red} What does this next part mean?} Now, over $\widetilde\Dcal$ there is a natural section of the projective bundle $\PP(F) \to \widetilde\Dcal$ (obtained by projectivising the diagonal embedding). If we therefore lift the above morphism to the projectivisation, we obtain a composition: 
\begin{equation*} \OO_{\PP(F)}(-1) \to \pi^\st F \to \pi^\st(\ev_q^\st \TT_H (\TT_{C_0})).\end{equation*}
The vanishing locus of this section is given by an Euler class calculation on $\PP(F)$. In order to obtain $\Dcal$ we must then intersect this vanishing locus with $\OO_{\PP(F)}(1)$. %(which represents the natural section $\widetilde\Dcal \to \PP(F)$.

Finally, we note that the class of $F$ {\bf \color{red} What class of $F$?} is computable in terms of data pulled back from the base. Consider the following bundle on $\Ecal$:
\begin{equation*} E = \bigoplus_{i=1}^r \left( \TT_{q_i}C_i \otimes \TT_{q_i}C_0\right). \end{equation*}
We may then identify $F$ with the pullback of $E$ twisted by the exceptional loci.

{\bf \color{red} More needs to be said here.}
\end{comment}

The main features of the discussion above are visible in the following example, which brings out the interplay between logarithmic moduli and blow-ups.

\begin{example}
Consider the following ray $\tau$ and sublocus $\rho$:
\begin{figure}[h]

@@

To establish a more direct connection with the geometry of the fibre product \eqref{IIIa fibre product}, recall that $\widetilde\Ecal$ has been built as an explicit modification of a finite cover of \eqref{IIIa fibre product}. For every $v\in\{\sqC_1,\ldots,\sqC_r\}$ - with the convention established at the end of Lemma \ref{lem:generic_proj_bundle} -, on $\widetilde\Ecal$ it makes sense to consider the line bundle $\OO(\delta_v)$ associated to a shortest path from the core to the circle in the $v$-direction. Construct $\mathcal P=\PP_{\widetilde\Ecal}(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))$; since $\sqC_1,\ldots,\sqC_r$ are always ordered on $\widetilde \Dcal$ (they may lie on the circle, be the first component off the circle, or degenerate and lie in the strict interior), there is a morphism $\widetilde \Dcal\to\mathcal P$ lifting $\widetilde \Dcal\to\widetilde\Ecal$. Consider the map of vector bundles on $\mathcal P$ (analogous to \eqref{eq:factorisation}):
\[ \OO_{\mathcal P}(-1)\to p^*(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))\xrightarrow{\sum\operatorname{d}f}\ev_{C_0}^*T_H.\]

>>>>

\subsubsection{$\Dcal$ from $\widetilde\Dcal$: first way}\label{subsection D from Dtilde} On $\widetilde\Dcal$ there is a tautological line bundle arising from the modular description of $\widetilde\VZ_{1,\alpha}(\PP^N|H,d)$, namely $\OO(\delta)$.

\begin{lemma}\label{lemma fibres of Odelta}
The fibre of $\OO(\delta)$ over a moduli point is naturally isomorphic to
\begin{equation*} T_{p}C^\prime \otimes T C_0\end{equation*}
where $C^\prime$ is any component of $C$ which lies on the radius, $p$ is the adjacent node which points towards the core, and $\TT_{C_0}$ is the universal tangent line bundle on the moduli space for $\sqC_0$.
\end{lemma}

\begin{proof}
Choose a component $C^\prime$ of the curve which lies on the circle. Then $\OO(\delta)$ is equal to the tensor product of tangent line bundles along the path of edges connecting $C^\prime$ to the core. However, this tensor product of line bundles is telescoping: any vertex which appears in the middle of the path will have an incoming and outgoing node, and since the vertex corresponds to a smooth rational curve these tangent spaces are naturally dual to each other~\cite[\S 2.2]{VZ}. Thus these factors cancel, and the result follows.
%Note that the final term on the right hand side is not given by $\TT_{q_i} C_0$ (where $q_i$ is the appropriate splitting node) but rather the tangent space to the core. 
\end{proof}
The content of Lemma \ref{lemma fibres of Odelta} is that the choice of point in the fibre of $\widetilde\Dcal \to \widetilde\Ecal$ provides a natural identification between the inward-pointing tangent spaces on the circle. This identification amounts to a specific linear dependence relation between the tangent vectors, which gives a point in the moduli space of attaching data for the corresponding elliptic singularity \cite[\S 2.2]{SMY2}. We may therefore express the factorisation condition in terms of the following natural morphism of vector bundles on $\widetilde\Dcal$:
\begin{equation*} \Sigma \operatorname{d}\!f \colon \OO(\delta)\otimes TC_0^\vee \to \ev_q^\st T H. \end{equation*}
Here $\Sigma \operatorname{d}\!f$ is defined by summing the images of the tangent vectors at the inward-pointing nodes on the circle, using the natural identification of each of these tangent spaces with $\OO(\delta)\otimes TC_0^\vee$ established above. This gives a section of the vector bundle
\begin{equation*} V = (\ev_q^\st TH) \otimes (TC_0) \otimes \OO(-\delta) \end{equation*}
which is transverse and whose vanishing locus coincides with $\Dcal$. Thus we obtain:
\begin{prop} \label{class of D} $[\Dcal] = \operatorname{e}(V) \cap [\widetilde\Dcal]$.\end{prop}
We claim that the class $\operatorname{e}(V)$ is tautological and computable. The only part which needs justification is $\OO(\delta)$. We claim that this can be obtained by pulling back a line bundle from $\Ecal$ and twisting by exceptional divisors.
\begin{lemma} For $i\in \{1,\ldots,r\}$ consider the line bundle
\begin{equation*} T_i = T_{q_i} C_i \otimes T_{q_i} C_0 \end{equation*}
on $\Ecal$, where $q_i$ is the splitting node. Then $\OO(\delta)$ is isomorphic to the pull-back of $T_i$ to $\widetilde\Dcal$ twisted by all exceptional divisors.\end{lemma}
We illustrate this lemma in the following example, which brings out the interplay between logarithmic moduli and blow-ups.
\begin{example}
Consider the following ray $\tau$ and sublocus $\rho$:
\begin{figure}[h]

@@

\subsubsection{$\Dcal$ from $\widetilde\Dcal$: second way} To establish a more direct connection with the geometry of the fibre product \eqref{IIIa fibre product}, recall that $\widetilde\Ecal$ has been built as an explicit modification of a finite cover of \eqref{IIIa fibre product}. For every $v\in\{\sqC_1,\ldots,\sqC_r\}$ - with the convention established at the end of Lemma \ref{lem:generic_proj_bundle} -, on $\widetilde\Ecal$ it makes sense to consider the line bundle $\OO(\delta_v)$ associated to a shortest path from the core to the circle in the $v$-direction. Construct $\mathcal P=\PP_{\widetilde\Ecal}(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))$; since $\sqC_1,\ldots,\sqC_r$ are always ordered on $\widetilde \Dcal$ (they may lie on the circle, be the first component off the circle, or degenerate and lie in the strict interior), there is a morphism $\widetilde \Dcal\to\mathcal P$ lifting $\widetilde \Dcal\to\widetilde\Ecal$. Consider the map of vector bundles on $\mathcal P$ (analogous to \eqref{eq:factorisation}):
\[ \OO_{\mathcal P}(-1)\to p^*(\bigoplus_{v\in \{\sqC_1,\ldots,\sqC_r\}}\OO(\delta_v))\xrightarrow{\sum\operatorname{d}f}\ev_{C_0}^*T_H.\]
